This journey has been possible due to so many helping hands in my life. 
First and foremost, my advisor Prof. Jignesh Patel. 
Jignesh's infectious energy and his passion for database research have shaped many ideas in my work.
I find it fascinating that he can dive into low level implementation details with the same ease as he can paint the big picture.
I benefited tremendously by his feedback on my writing and presentation skills. 
He always prioritized my well being and my family's well being over other things. 
I hope to emulate his industriousness, discipline and remarkable interpersonal communication skills. 

I was fortunate to be in Wisconsin's database group and interact with some astounding professors. 
Jeff Naughton during his time in Wisconsin was like a father figure. 
His sharp and witty comments during the talks always made the atmosphere lively. 
I hope to enjoy Jeff's mentorship in my professional life. 
I have had numerous interactions with AnHai Doan.
I am grateful for his advice and support during my job search.
I thank my committee members Professors Paris Koutris, Theodoros Rekatsinas, Andrea Arpaci-Dusseau and Remzi Arpaci-Dusseau for reading my dissertation and for their helpful comments. 

I am grateful to the Computer Sciences Graduate Coordinator Angela Thorp for her support in administrative matters. 

I could live a debt free grad school life, thanks to the grants from the National Science Foundation.
This money essentially comes from the tax dollars of hardworking American men and women.
My sincere thanks to all of them.

Quickstep is a group project and is built from the contributions of several students.
My work would not have been possible without their collective efforts. 
I would like to thank Quickstep developers including Craig Chasseur, Qiang Zeng, Shoban Chandrabose, Yinan Li, Gerald, Zuyu Zhang, Jianqiao Zhu, Rathijit Sen, Saket Saurabh, Navneet Potti, Udip Pant, and Quickstep's Apache mentor Julian Hyde.  

I got a chance to do multiple internships in grad school. 
I am grateful to my Samsung internship colleagues Chang-Ho Choi and Suraj Waghulde for their support. 
I learnt a great deal at my internship with Microsoft CISL.
Avrilia Floratou was a great mentor and I learnt from her the \textit{modus operandi} of industrial research.
Ashvin Agrawal was a great collaborator who taught me a great deal about rapid software prototyping. 
Prajakta Kalmegh and I became good friends through car pool. 
%The really smart and nice people in CISL including Sriram Rao, Carlo Curino, Subramaniam Krishnan, Chris Cai, Konstantinos Karanasos, Chris Douglas, Virajith Jalaparty, and Arun Suresh made the internship a memorable experience. 

Life would have been very difficult without some great friends in the database group. 
Yash Govind has been a great friend and confidant.
Shaleen Deep was the sounding board for a lot of my silly ideas.
Adalbert Gerald Soosai Raj has provided immense support through good times and bad. 
I developed close friendship with Hakan Memisoglu during his time in Wisconsin.
Whenever I needed help, Rogers Jeffrey Leo John would always come along with his (poor?) jokes. 
I learnt a great deal from the long interactions with Bruhathi on the pipelining project, as well as during my job search.
Pradap Konda Venkatraman's mature demeanor and deep insights had a calming effect on me.
I thank Adel Ardalan for his help on my various presentations. 
Paul Suganthan GC helped me find accommodation during my first summer in Wisconsin. 
I had a great time working with Vinitha Reddy on many course projects.

Through the years, I had the company of some wonderful office mates. 
They often accompanied me to Peet's for coffee breaks and helped me refresh from the work.
The long chats with Hakan Memisoglu and Marc Spehlmann made walks to Gordon dining center seem like a breeze.
I am grateful to Spyros Blanas for his tidbits of wisdom. 
An incomplete list of other officemates include Sriram Sridharan, Yueh-Hsuan Chiang, Lalitha Viswanathan, Han Li, and Dylan Bacon.

I got to know several bright and kind people in the UW over the years. 
Prashant Saxena was my roommate for the first two years and his sense of humor would always lighten up the atmosphere. 
I had several insightful discussions with Prashant over the delicious dinners he cooked. 
I thoroughly relished Ankit Agrawal's company during his Master's.
Ankit and I could talk for hours about anything under the sun.
I still remember Friday evening tea parties with Ankit and Nishant Agrawal. 
Sanketh Nalli remains a good friend till date.
Meenakshi Syamkumar has become a close family friend. 
I am grateful to Ramnatthan Alagappan and Aishwarya Ganesan for their company for dinners and fun outings.
An incomplete list of friends in the UW include Rohit Bhat, Junaid Khalid, Ashutosh Pandey, Swapnil Haria, Amrita Roy Chowdhury, Neha Godwal, Kushagra Garg, and Sakshi Gupta. 

My under-graduation institute BITS Pilani has a special place in my life.
Many friendships from BITS have accompanied me from India to the US. 
I have had many trips in grad school with these friends which helped me refresh.

Soumyadeep Ghosh has been a close friend ever since under-graduation days and till date he's my goto person for venting out frustration. 
I am grateful to Aditya Vijay's great advice in so many situations.
Ankur Gupta provided great company on various trips that we went on together.
%My friends from Pilani continue to provide me moments of great joy.
Ameya Zambre was the first familiar face I saw after entering the USA. 
I am greateful to Nilesh Deshpande, Jhinak Sen, Vineet Pandey, Apoorv Bapat, Akash Badave, Aditya Belsare, Aditya Sharma, Ashirvad Singh, Chanchal Batra, Abhishek Jain, Subhro Sarkar, Priyanka Gupta and Pooja Chandra.

%I spent the first two years of my grad school working on the BITS2MSPhD project to provide MS/PhD related assistance to aspiring undergraduates from BITS. 
%Volunteering for BITS2MSPhD while in grad school helped me manage my time better.
%I had a great time working with Varsha Rajshekhar, Yash Gandhi, Manish Kumar, Adwait Gandhe and Ishaan Biswas. 

Living in Madison would be dull without a fantastic group of friends.
I met a lot of people through Maharashtra Mandal in Madison.
Thanks to all of them for the joyful cultural events and parties.
%I am grateful to Kimaya and Rohan Pradhan, Girish and Priti Kulkarni, Nikhil and Tanuja Jog, Abhishek and Nishi Kulkarni, Milind and Rama Ranade, Abhinav and Nishita Verma, Stimit and Vidhi Shah, and Ojas Kale for the numerous parties and fun times together. 

Family is the ultimate support system. 
My in laws Anant and Neha Hirurkar always supported my pursuit of PhD. 
My sisters Sukhada and Sampada and their husbands Akshay and Hitesh encouraged me to follow my aspirations and took great care of my parents.
My nephews and niece provided great joy during my trips to India.

I dearly miss my late maternal grandmother. 
She was a single mother and a primary school teacher.
She raised six kids on a meager income.
Despite the hardships, she had a staunch belief in the power of education.
She has a profound influence on my life.

I can probably never do enough to repay what my parents have done for me. 
They made several sacrifices to ensure that I got a high quality education. 
They always believed in my decisions and supported me in tough times.
Whatever little success that I have achieved is a testimony to their strong foundational values.
Thank you Aai and Baba from the bottom of my heart.

I got the gift of fatherhood during grad school.
Being a father to an adorable little girl felt daunting at time, but ultimately was a very rewarding feeling.
All the stress from working in grad school would go away with a single smile of my daughter.

Last, but by no means the least, I would like to thank my wonderful wife Supriya. 
She always believed in me, even when I doubted myself. 
In bad times or good, she always kept assuring me that I was capable of making it to the finish line. 
She went to such extremes that I was not allowed to crack jokes on myself! 
She sacrificed her professional life and single handedly managed our family life so that I could focus on my PhD. 
Supriya, you made all the hardships worthwhile! 